------------------------------------------------------------------------
title: R_Basics
author: Derek Ding 
date: "`r Sys.Date()`" 
output: html_document
code_folding: show
------------------------------------------------------------------------
## 1. Functions

__Key Points:__

__1.1__ Define a function using name <- function(...args...) {...body...}.

__1.2__ Call a function using name(...values...).

__1.3__ R looks for variables in the current stack frame before looking for them at the top level.

__1.4__ Use help(thing) to view help for something.

__1.5__ Put comments at the beginning of functions to provide help for that function.

__1.6__ Annotate your code!

__1.7__ Specify default values for arguments when defining a function using name = value in the argument list.

__1.8__ Arguments can be passed by matching based on name, by position, or by omitting them (in which case the default value is used).


__1.9__ Add ? and ?? before the function to search its documentations. 

```{r echo= TRUE, error=TRUE}

example.func <- function(arg_1, arg_2) {
   result.sum = arg_1 + arg_2
  return(result.sum)
}

example.func(2,5)

#?example.func
#??example.func

```



## 2. Arguments
Functions like Sys.Date() takes no arguments, 

Some like factorial(4) takes a single argument.

Others takes multiple arguments, like paste(). In which case, sep = " " is the
default value, unless otherwise specify. sep here is an optional argument.

args(function): show what the arguments are.

Best practice is to specify clear, the value for each arg in R:
example.func(arg_1 = 2, arg_2 = 9)

```{r echo= TRUE, error=TRUE}
Sys.Date()
factorial(4)
paste("Hello", "World!", sep=" ")
paste("Hello", "World!")
paste("Hello", "World!", "World!", sep=" +_+_+_+_+_:D ")

args(example.func)

example.func <- function(arg_1, arg_2 = 8) {
   result.sum = arg_1 + arg_2
  return(result.sum)
}

example.func(arg_1 = 2)
example.func(arg_1 = 2, arg_2 = 9)
example.func(3,9)

```

## 3. Obeject
Storing the values.

```{r echo= TRUE, error=TRUE}
num1 <- log(8)
num2 <- 0.5

num1
num2
example.func(num1, num2)
```

## 4. Vectors

r stores values in vectors, initialize a vector with c()

What does Rnorm in R do?
rnorm is the R function that simulates random variates having a specified normal 
distribution. As with pnorm , qnorm , and dnorm , optional arguments specify the 
mean and standard deviation of the distribution.

```{r echo= TRUE, error=TRUE}
rnorm(18)
```

For example, here pmax function goes from vec1(1) vs. vec2(1), and vec1(2) vs. vec2(2), and so on.
That is the general mechanics of R multiple vectors functions.
```{r echo= TRUE, error=TRUE}
sample.vector <- c(1, 2, 3, 4, 15)
factorial(sample.vector)
mean(sample.vector)
sample.vector2 <- c(6, 7, 8, 9, 10)
pmax(sample.vector, sample.vector2)
```


## 5. Data Types
R recognizes 6 different types of data and most of the default functions can 
operate on different types of data pretty intuitively.
-   logical: **TRUE** and **FALSE**
-   numeric: 3, 157.5
-   integer: 1L, 6L...
-   complex: 6 + 5i complex numbers with real and imaginary parts.
-   character: "a", "apple"
-   raw: Coercion to raw treats the input values as representing small (decimal) 
integers, so the input is first coerced to integer, and then values which are 
outside the range [0 ... 255] or are NA are set to 0 (the nul byte).

as.raw and is.raw

```{r echo= TRUE, error=TRUE}
num.vec = c(1, 2, 7)
str.vec = c("apple","orange","pear")

num.vec + 5
nchar(str.vec)

# numeric vectors:
typeof(num.vec)

# characters vectors:
typeof(str.vec)

# integer vectors:
typeof(c(1L,11L,111L,22L,9L))

# logical vectors
typeof(c(TRUE, FALSE, TRUE))

# complex
typeof(c(6+5i, 7+8i, 0+9i))

# raw
typeof(as.raw(40))

# by default, R will collapse multiple types into char type if specified in one
# vector
typeof(c(1, FALSE, "hello"))

# more on raw
as.raw(40)
charToRaw("A9") # seems to be more of a ASCII converting function

```


## 6. Lists
In addition to vectors, R comes with another data-structure: list.
List is a vector of vectors, each vector stores a single type of data, but a list
can store a various types of data, each element in a list is another vector on
its own.

You can also name each element and use the dollar sign to call each element by its
name.

```{r echo= TRUE, error=TRUE}
a_list <- list(
  1:5, "apple", c("apple","orange","pear"), TRUE)
a_list

a.list.2 <- list(
  numbers=1:9, # automatically initialized as int
  fruits = c("apple","orange","pear"),
  logicals = c(TRUE, TRUE, FALSE)
  )

a.list.2
typeof(a.list.2)
a.list.2$fruits
typeof(a.list.2$logicals)
typeof(a.list.2$numbers)
typeof(a.list.2$fruits)
```

## 7. Packages

Base R vs. Packages.

e.g. ggplot function comes with ggplot2 package, we won't be able to use ggplot
if we do not load the ggplott2 package.

```{r echo= TRUE, error=TRUE}
install.packages(c("ggplot2", "forecast"))
```

